\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Tweaks to appendix

There is no particular reason, however, that ARGs must be
retrospective. They can also be \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell; see Appendix XXX),
and are annotated
with genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix XXX)
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which nodes can be unary in the local trees
of an ARG. Furthermore, ARGs inferred by current methods contain
locally unary nodes from several of them (see section XXX).

@@

@@

The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_10$
(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
we can only represent this as a instantanous trifucation or polytomy.

There is biological significance to this cellular-level inheritance process. Were a mutation to occur
during the first cell division of $D_10$, the two gametes produced by the cells
in the left half of $D_10$ could share a mutation not present in the right hand gamete. In other words,
the relationship between three (or more) gametes that originate, without recombination, from the same parental
chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
relationships will be impossible to infer from the data, with subsequent implications for
using the eARG representation to store the results of inference.

As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
by a polytomy in the local trees.
\begin{figure}
\begin{center}
    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
\end{center}
\caption{\label{fig-cell-lines}
Cellular inheritance of a single chromosome in a diploid population.
Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
meiosis is not illustrated in detail.
Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
think solid line and involves a single recombination event and four coalescence events
(highlighted as deep red and blue squares within individuals $D_5$, $D_10$, and $D_13$).
ARG lineages also show gametic genomes, contained within shaded circles.
As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
shaded by the number of descendant samples.
}
\end{figure}

>>>>

There is no particular reason, however, that ARGs must be
retrospective. They can also be \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations
% Wilder found cell-to-cell a little confusing here, we could delete
% if necessary (it's tackled later on)
(ultimately from cell to cell; see Appendix~\ref{cell-lineages-and-args}),
and are annotated
with genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which nodes can be unary in the local trees
of an ARG. Furthermore, ARGs inferred by current methods contain
locally unary nodes from several of them (see section XXX).

@@

\begin{figure}
\begin{center}
    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
\end{center}
\caption{\label{fig-cell-lines}
Cellular inheritance of a single chromosome in a diploid population.
Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
meiosis is not illustrated in detail.
Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
think solid line and involves a single recombination event and four coalescence events
(highlighted as deep red and blue squares within individuals $D_5$, $D_{10}$, and $D_{13}$).
ARG lineages also show gametic genomes, contained within shaded circles.
As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
shaded by the number of descendant samples.
}
\end{figure}

@@

The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_{10}$
(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
we can only represent this as a instantaneous trifurcation or polytomy.

There is biological significance to this cellular-level inheritance process. Were a mutation to occur
during the first cell division of $D_{10}$, the two gametes produced by the cells
in the left half of $D_{10}$ could share a mutation not present in the right hand gamete. In other words,
the relationship between three (or more) gametes that originate, without recombination, from the same parental
chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
relationships will be impossible to infer from the data.

As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
by a polytomy in the local trees.
